Formulating the posterior for \tardis

### 2015-08-20 posterior/model.tex
[posterior] Fix spelling

\subsubsection*{What we want}
The posterior distribution of $\vecth$ given the data $\data$,
$P(\vecth | \data)$, is given by  Bayes' theorem as
\begin{equation}
  \label{eq:Bayes-full}
  P\given{\vecth}{\data} = \frac{P\given{\data}{\vecth} P(\vecth)}{\int \rmdx{\vecth} P\given{\data}{\vecth} P(\vecth)} \,.
\end{equation}
Technically, we should write every propability (density) as
conditional on a specific statistical model $M$,
e.g. $P\given{\vecth}{\data, M}$.  At present, we are not interested
in comparing different models but rather in extracting the parameters
of one model. Thus we omit $M$ and can ignore the denominator (the
\emph{evidence}) in Bayes' theorem. All we need is to define the
likelihood $\likelihood(\vecth) \equiv P\given{\data}{\vecth}$ and the
prior $P(\vecth)$ in
\begin{equation}
  \label{eq:Bayes-simple}
  P\given{\vecth}{\data} \propto \likelihood(\vecth) P(\vecth) \,.
\end{equation}

@@

For virtual packets, the luminosity distribution seems to depend
mildly on the frequency.  We do not know $P(\Lumi | 1, \vecth)$ before
we do the simulation and we do not know its functional form
either. But we want to minimize the assumptions on the form. Hence
seek a nonparametric description of the samples. Obviously, if we
observe $n_i=0$ or $n_i=1$, we cannot infer the distribution $P(\Lumi
| 1, \vecth)$ from such little information. If we may assume that
$P(\Lumi | 1, \vecth)$ varies only little with $i$, we may pool
together luminosities $l_{ij}, j=1 \dots n_i$, of adjacent bins to
form a handful of larger bins, say $10-100$, in which we assume the
distribution is constant and we have enough packets, at least $\sum_i
n_i \gtrsim 500$. The I propose we use Gaussian kernel density
estimation (KDE) to obtain a smooth estimate of $P(\Lumi | 1, \vecth)$
that we can evaluate at arbitray $\Lumi$. In python, the easiest way
to do this is via \texttt{scipy.stats.gaussian\_kde}. The fastest to
my knowledge is \texttt{figtree} but the difference will be mostly
irrelevant unless there $\sum_i n_i = 10^5$ packets in a combination
of bins.

Given the KDE estimate, we could compute the distribution for the sum
of two variables as the convolution
\begin{align}
  \label{eq:sum-of-2}
  P(\Lumi | N_i=2, \vecth) = \int \rmdx{l} P(\Lumi -l | N_i=1, \vecth) P(l | N_i=1, \vecth).
\end{align}
For $N_i =k$, we would repeat the procedure convolving $P(\cdot |
N_i=k-1, \vecth)$ and $P(\Lumi | N_i=1, \vecth)$ and each time we
would need a new KDE estimate and we would have to repeat the integral
for different values of $\Lumi$. This quickly gets out of hand and I
fear the errors due to KDE assumptions add up. In my attempts with
\texttt{mathematica}, the probability for many values of $\Lumi$ was
negative and the whole procedure is slow. Let's forget about it.

@@

For $N \gtrsim 15$, we can just use the central limit theorem
such that
\begin{align}
  \label{eq:clt}
  P(\Lumi | N_i, \vecth) \approx \gaussian \given{\Lumi}{N_i \times E[l], N_i \times V[l]}
\end{align}
and the expecation value and variance have to be estimated by the mean
and variance of the samples $\{l_j\}$. In fact, this paper~\cite{bohm_statistics_2014}
shows that the \emph{scaled Poisson} is a more accurate approximation
given the first and second moment. But the resulting integral can't be
solved exactly, so for easier calculation we stick with the Gaussian.

@@

For $n_i=0$, the distribution diverges at zero, a consequence of the
Jeffreys prior.  For $n_i \gtrsim 30$, the distribution can be well
approximated by a Gaussian
\begin{align}
  \label{eq:poisson-gauss-approx}
  P(N_i | \vecth) \approx \gaussian \given{N_i}{n_i-1, 2 n_i + 1}
\end{align}

@@

The strategy would be to look at each bin separately. If $n_i \gtrsim
5$, use the large-$n_i$ approach from above. Else start at $N_i=1$ and
sum the numerical integrals until it is $10^{-4}$ of the maximum
term. For 10000 bins, this works can be parallelized easily.

For the real packets, I compared the general approach to the Gaussian
large-$n_i$ approximation, using numerical integration in both cases
in \texttt{mathematica}. For $n_i=1\dots5$, the difference is slighly
less than 10\%. From $n_i=10$ to 15, it drops from 5\% to 3\% level,
and for $n_i=100$, both approaches agree to within $10^{-4}$. Laplace
can be used from $n_i=5$ on. In all cases except $n_i=100$, the
Gaussian approximation is systematically \emph{below} the general
result.

If we are ok with less than 10\% error, we can employ the Gaussian
approximation from $n_i\ge5$, which means we neglect the 3rd and
higher moments of the luminosity distribution. If the virtual packets
are even more skewed than the real packets, we might need to start at
$n_i \sim 10$.

@@

Now suppose we want to include the available prior information. The
actual numbers come from table 5.1 of Wolfgang's thesis. He lists
constraints of the type
\begin{equation}
  \begin{aligned}[t]
    \label{eq:abundance-constraints}
    \mbox{C}  &< 12.5 \%,\\
    \mbox{Si} &>1 \%,\\
    1 &<\mbox{Si/S ratio} < 10 .
  \end{aligned}
\end{equation}
The most straightforward approach is to incorporate them as hard cuts
into the prior, leading to a new normalization. We will have to refine
this strategy if the regions of high posterior probability turn out to
be very close to the cuts. Formally, suppose $V$ is the subvolume of
the unit simplex in which all constraints are satistfied. Outside of
$V$, we set $P(\vecth) \equiv 0$. In the code, a sequence of
\texttt{if} statements incorporates the constraints. If any is
violated, the prior on the log scale is $-\infty$ and no tardis
simulation needs to be run. Our prior thus becomes
\begin{equation}
  \label{eq:abundances-prior}
  P(\vecth) = C_V^{-1} \mathbf{1}_V(\vecth) \Dirichlet\left(\vecmu(\vecth) \middle| \vecalpha=(10^{-4}, \dots, 10^{-4})\right) \boldsymbol{\theta}
\end{equation}
where the normalization constant $C_V$ is  determined from
\begin{equation}
  \begin{aligned}
    \label{eq:abundance-normalization}
    1 &= \int \rmdx{\vecth} P(\vecth)\\
    \Rightarrow C_V  &= \int_V \rmdx{\vecth} \Dirichlet\left(\vecmu(\vecth) \middle| \vecalpha=(10^{-4}, \dots, 10^{-4})\right)
  \end{aligned}
\end{equation}
and $\mathbf{1}_V$ is the indicator function on $V$; i.e.,
$\mathbf{1}_V(\vecth) = 1$ iff $\vecth \in V$.  If we keep the cuts
fixed, we only need to determine $C_V$ once. We can do that with good
precision without nested sampling by using the law of large numbers as
\begin{equation}
  \label{eq:abundances-large-numbers}
  C_V \approx \frac{1}{N} \sum_{i=1}^N \mathbf{1}_V(\vecth), \; \vecth \sim \Dirichlet\left(\vecmu(\vecth) \middle| \vecalpha=(10^{-4}, \dots, 10^{-4})\right)
\end{equation}
and drawing directly from $\Dirichlet(\vecalpha)$ using
\texttt{scipy.random.dirichlet}. However, we don't really need to
normalize our densities properly unless we compare different
models. In the near future, we will only infer the parameters of a
single model so there is no need to compute $C_V$.

>>>>

\subsubsection*{What we want}
The posterior distribution of $\vecth$ given the data $\data$,
$P(\vecth | \data)$, is given by  Bayes' theorem as
\begin{equation}
  \label{eq:Bayes-full}
  P\given{\vecth}{\data} = \frac{P\given{\data}{\vecth} P(\vecth)}{\int \rmdx{\vecth} P\given{\data}{\vecth} P(\vecth)} \,.
\end{equation}
Technically, we should write every probability (density) as
conditional on a specific statistical model $M$,
e.g. $P\given{\vecth}{\data, M}$.  At present, we are not interested
in comparing different models but rather in extracting the parameters
of one model. Thus we omit $M$ and can ignore the denominator (the
\emph{evidence}) in Bayes' theorem. All we need is to define the
likelihood $\likelihood(\vecth) \equiv P\given{\data}{\vecth}$ and the
prior $P(\vecth)$ in
\begin{equation}
  \label{eq:Bayes-simple}
  P\given{\vecth}{\data} \propto \likelihood(\vecth) P(\vecth) \,.
\end{equation}

@@

For virtual packets, the luminosity distribution seems to depend
mildly on the frequency.  We do not know $P(\Lumi | 1, \vecth)$ before
we do the simulation and we do not know its functional form
either. But we want to minimize the assumptions on the form. Hence we
seek a nonparametric description of the samples. Obviously, if we
observe $n_i=0$ or $n_i=1$, we cannot infer the distribution $P(\Lumi
| 1, \vecth)$ from such little information. If we may assume that
$P(\Lumi | 1, \vecth)$ varies only little with $i$, we may pool
together luminosities $l_{ij}, j=1 \dots n_i$, of adjacent bins to
form a handful of larger bins, say $10-100$, in which we assume the
distribution is constant and we have enough packets, at least $\sum_i
n_i \gtrsim 500$. The I propose we use Gaussian kernel density
estimation (KDE) to obtain a smooth estimate of $P(\Lumi | 1, \vecth)$
that we can evaluate at arbitrary $\Lumi$. In python, the easiest way
to do this is via \texttt{scipy.stats.gaussian\_kde}. The fastest to
my knowledge is \texttt{figtree} but the difference will be mostly
irrelevant unless there $\sum_i n_i = 10^5$ packets in a combination
of bins.

Given the KDE estimate, we could compute the distribution for the sum
of two variables as the convolution
\begin{align}
  \label{eq:sum-of-2}
  P(\Lumi | N_i=2, \vecth) = \int \rmdx{l} P(\Lumi -l | N_i=1, \vecth) P(l | N_i=1, \vecth).
\end{align}
For $N_i =k$, we would repeat the procedure convoluting $P(\cdot |
N_i=k-1, \vecth)$ and $P(\Lumi | N_i=1, \vecth)$ and each time we
would need a new KDE estimate and we would have to repeat the integral
for different values of $\Lumi$. This quickly gets out of hand and I
fear the errors due to KDE assumptions add up. In my attempts with
\texttt{mathematica}, the probability for many values of $\Lumi$ was
negative and the whole procedure is slow. Let's forget about it.

@@

For $N \gtrsim 15$, we can just use the central limit theorem
such that
\begin{align}
  \label{eq:clt}
  P(\Lumi | N_i, \vecth) \approx \gaussian \given{\Lumi}{N_i \times E[l], N_i \times V[l]}
\end{align}
and the expectation value and variance have to be estimated by the mean
and variance of the samples $\{l_j\}$. In fact, this paper~\cite{bohm_statistics_2014}
shows that the \emph{scaled Poisson} is a more accurate approximation
given the first and second moment. But the resulting integral can't be
solved exactly, so for easier calculation we stick with the Gaussian.

@@

For $n_i=0$, the distribution peaks at zero at a value of $70\%$, a
number sensitive to the choice of prior $P(\lambda_i)$.  For $n_i \gtrsim 30$, the
distribution can be well approximated by a Gaussian
\begin{align}
  \label{eq:poisson-gauss-approx}
  P(N_i | \vecth) \approx \gaussian \given{N_i}{n_i-1, 2 n_i + 1}
\end{align}

@@

The strategy would be to look at each bin separately. If $n_i \gtrsim
5$, use the large-$n_i$ approach from above. Else start at $N_i=1$ and
sum the numerical integrals until it is $10^{-4}$ of the maximum
term. For 10000 bins, this work can be parallelized easily. It's all
elementary function, so this can be done in \texttt{C} without too
much effort.

For the real packets, I compared the general approach to the Gaussian
large-$n_i$ approximation, using numerical integration in both cases
in \texttt{mathematica}. For $n_i=1\dots5$, the difference is slightly
less than 10\%. From $n_i=10$ to 15, it drops from 5\% to 3\% level,
and for $n_i=100$, both approaches agree to within $10^{-4}$. Laplace
can be used from $n_i=5$ on. In all cases except $n_i=100$, the
Gaussian approximation is systematically \emph{below} the general
result.

If we are OK with less than 10\% error, we can employ the Gaussian
approximation from $n_i\ge5$, which means we neglect the 3rd and
higher moments of the luminosity distribution. If the virtual packets
are even more skewed than the real packets, we might need to start at
$n_i \sim 10$.

@@

Now suppose we want to include the available prior information. The
actual numbers come from table 5.1 of Wolfgang's thesis. He lists
constraints of the type
\begin{equation}
  \begin{aligned}[t]
    \label{eq:abundance-constraints}
    \mbox{C}  &< 12.5 \%,\\
    \mbox{Si} &>1 \%,\\
    1 &<\mbox{Si/S ratio} < 10 .
  \end{aligned}
\end{equation}
The most straightforward approach is to incorporate them as hard cuts
into the prior, leading to a new normalization. We will have to refine
this strategy if the regions of high posterior probability turn out to
be very close to the cuts. Formally, suppose $V$ is the subvolume of
the unit simplex in which all constraints are satisfied. Outside of
$V$, we set $P(\vecth) \equiv 0$. In the code, a sequence of
\texttt{if} statements incorporates the constraints. If any is
violated, the prior on the log scale is $-\infty$ and no tardis
simulation needs to be run. Our prior thus becomes
\begin{equation}
  \label{eq:abundances-prior}
  P(\vecth) = C_V^{-1} \mathbf{1}_V(\vecth) \Dirichlet\left(\vecmu(\vecth) \middle| \vecalpha=(10^{-4}, \dots, 10^{-4})\right) \boldsymbol{\theta}
\end{equation}
where the normalization constant $C_V$ is  determined from
\begin{equation}
  \begin{aligned}
    \label{eq:abundance-normalization}
    1 &= \int \rmdx{\vecth} P(\vecth)\\
    \Rightarrow C_V  &= \int_V \rmdx{\vecth} \Dirichlet\left(\vecmu(\vecth) \middle| \vecalpha=(10^{-4}, \dots, 10^{-4})\right)
  \end{aligned}
\end{equation}
and $\mathbf{1}_V$ is the indicator function on $V$; i.e.,
$\mathbf{1}_V(\vecth) = 1$ iff $\vecth \in V$.  If we keep the cuts
fixed, we only need to determine $C_V$ once. We can do that with good
precision without nested sampling by using the law of large numbers as
\begin{equation}
  \label{eq:abundances-large-numbers}
  C_V \approx \frac{1}{N} \sum_{i=1}^N \mathbf{1}_V(\vecth), \; \vecth \sim \Dirichlet\left(\vecmu(\vecth) \middle| \vecalpha=(10^{-4}, \dots, 10^{-4})\right)
\end{equation}
and drawing directly from $\Dirichlet(\vecalpha)$ using
\texttt{scipy.random.dirichlet}. However, we don't really need to
normalize our densities properly unless we compare different
models. In the near future, we will only infer the parameters of a
single model so there is no need to compute $C_V$.
